Emergent dual scaling of riverine biodiversity

### 2021-03-24 manuscript_pnas.Rmd
edits: maintext, emphasize scaling law

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions (Hokkaido Island in Japan and the midwestern United States) to show that ecosystem size and complexity play equally-important roles in scaling biodiversity in rivers. The recurrent merging of various streams and rivers forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks supported greater species richness due to increased space and environmental heterogeneity. Both relationships were linear on a log-log scale, indicating a general power-law scaling along the two orthogonal dimensions of ecosystem structure, i.e., size and complexity. In support of this theoretical prediction, the dual power-law scaling of riverine fish diversity has consistently emerged across the study regions despite hosting different fauna with distinct evolutionary histories. The dual scaling of biodiversity may be a pervasive emergent property of branching networks with important implications for biodiversity conservation.

@@

Here, we show that ecosystem size and complexity play comparable roles in scaling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

@@

Our theoretical analysis yielded results consistent with our prediction. Ecosystem size and complexity both increased $\gamma$ diversity under a realistic landscape scenario (**Figures 2 and 3**), where the environmental variation at headwaters $\sigma_h$ is greater than the degree of local environmental noise $\sigma_l$ (**Figure 1A**). The relationships had a characteristic of power-law (i.e., linear in a log-log scale) and were consistent under various ecological scenarios (compare panels in **Figures 2 and 3**). Importantly, the impact of ecosystem complexity was comparable to that of ecosystem size. Hence, regardless of ecological scenarios, ecosystem size and complexity are likely to be equally important in regulating $\gamma$ diversity.

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This pattern reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@leiboldMetacommunityConceptFramework2004; @thompsonProcessbasedMetacommunityFramework2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@leiboldMetacommunityConceptFramework2004; @economoSpeciesDiversityNeutral2008; @thompsonProcessbasedMetacommunityFramework2020]. Therefore, ecosystem size and complexity can increase $\gamma$ diversity via enhanced species sorting or mass effects, and which takes primacy is dependent on the degree of dispersal limitation. These patterns were qualitatively similar across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar patterns in $\gamma$ diversity emerge through different ecological pathways.

@@

We found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`) despite the substantial difference in fish fauna between the study regions (**Tables S5 and S6**). The estimated $\gamma$ diversity linearly increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) on a log-log scale (i.e., power-law), and these patterns were consistent across the study regions (**Table 1 and Figure 4**). Thus, watershed-scale riverine biodiversity likely obeys a general scaling law along the two orthogonal dimensions of ecosystem structure, i,e, size and complexity. In particular, the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supporting Information**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). This result is reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual scaling of $\gamma$ diversity to emerge. Further, the consistent patterns across the geographic regions with distinct evolutionary histories imply that ecological traits of species in metacommunities may have a minor effect on the scaling relationships, as predicted by our theory.

@@

Using theory and extensive community data, our study provides strong evidence that ecosystem size and complexity play equally important roles in shaping riverine biodiversity. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

>>>>

**Abstract:** A prevailing paradigm suggests that species richness increases with area in a decelerating way; this ubiquitous power-law scaling, the species-area relationship, has formed the foundation of many conservation strategies. In spatially complex ecosystems, however, the area may not be the sole dimension to scale biodiversity patterns because the scale-invariant complexity of fractal ecosystem structure may drive ecological dynamics in space. Here, we use theory and analysis of extensive fish community data from two distinct geographic regions (Hokkaido Island in Japan and the midwestern United States) to show the emergence of robust scaling law along the two orthogonal dimensions of ecosystem size and complexity in rivers. The recurrent merging of various streams and rivers forms fractal branching systems, where the prevalence of branching (ecosystem complexity) represents a macro-scale control of the ecosystem's habitat heterogeneity. In the meantime, ecosystem size dictates metacommunity size and total habitat diversity, two factors regulating biodiversity in nature. Our theory predicted that, regardless of simulated species' traits, larger and more branched 'complex' networks supported greater species richness due to increased space and environmental heterogeneity. Both relationships were linear on a log-log scale, indicating a robust scaling law along the dimensions of size and complexity. In support of this theoretical prediction, the dual power-law scaling of riverine fish diversity has consistently emerged across the study regions despite hosting different fauna with distinct evolutionary histories. The dual scaling of biodiversity may be a pervasive emergent property of branching networks with important implications for biodiversity conservation.

@@

Here, we show the emergence of robust scaling law along the two orthogonal dimensions of ecosystem size and complexity in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

@@

Our theoretical analysis yielded results consistent with our prediction. Ecosystem size and complexity both increased $\gamma$ diversity under a realistic landscape scenario (**Figures 2 and 3**), where the environmental variation at headwaters $\sigma_h$ is greater than the degree of local environmental noise $\sigma_l$ (**Figure 1A**). The power-law relationships (i.e., linear in a log-log scale) have emerged along the both axes of ecosystem size and complexity irrespective of ecological scenarios (compare panels in **Figures 2 and 3**). Importantly, the impact of ecosystem complexity was comparable to that of ecosystem size. Hence, regardless of ecological scenarios, ecosystem size and complexity are likely to be equally important in scaling $\gamma$ diversity.

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This pattern reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@leiboldMetacommunityConceptFramework2004; @thompsonProcessbasedMetacommunityFramework2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@leiboldMetacommunityConceptFramework2004; @economoSpeciesDiversityNeutral2008; @thompsonProcessbasedMetacommunityFramework2020]. Therefore, ecosystem size and complexity can increase $\gamma$ diversity via enhanced species sorting or mass effects, and which takes primacy is dependent on the degree of dispersal limitation. These patterns were qualitatively similar across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar scaling in $\gamma$ diversity emerge through different ecological pathways.

@@

We found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`) despite the substantial difference in fish fauna between the study regions (**Tables S5 and S6**). The estimated $\gamma$ diversity linearly increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) on a log-log scale (i.e., power-law), and the estimated scaling exponents were consistent across the study regions (`r op(coef(fit$gamma$model)[2])` for watershed area and `r op(coef(fit$gamma$model)[3])` for branching probability; **Table 1 and Figure 4**). Thus, watershed-scale riverine biodiversity likely obeys a general scaling law along the two orthogonal dimensions of ecosystem size and complexity. The effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supporting Information**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). This result is reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual scaling of $\gamma$ diversity to emerge. Further, the consistent patterns across the geographic regions with distinct evolutionary histories imply that ecological traits of species in metacommunities may have a minor effect on the scaling relationships, as predicted by our theory.

@@

Using theory and extensive community data, our study provides strong evidence that ecosystem size and complexity play equally important roles in shaping riverine biodiversity. In particular, the emergent power-law scaling between ecosystem complexity and biodiversity points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].
